---
title: 'Dissertation'
output: 
  html_document:
    toc: true
    number_sections: true
    code_folding: show
date: "2022-10-27"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning=FALSE) #doesn't show code or error messages in knit
```

```{r library, message=FALSE, include=FALSE} 
#don't print outputs
library(dplyr)
library(tidyverse) 
library(knitr)
library(readr)
library(ggplot2)
library(bbmle)
library(mizer) 
library(Hmisc)
library(lme4)
```

```{r block, message=FALSE, include=FALSE}
(load("stomach_dataset.Rdata")) #change to access via github repository
#the dataset is already named stom_df
stom_df
df <- stom_df %>% transmute(haul_id, latitude, longitude, ices_rectangle, year, pred_species, pred_weight_g, pred_length_cm, prey_species, prey_weight_g, prey_type = prey_funcgrp, indiv_prey_weight = prey_ind_weight_g, no._prey_per_stmch = prey_count / n_stomachs, ppmr) 
df
#ices_rectangle accounts for location of points
```

```{r cleaning smaller, message=FALSE, include=FALSE}
df_fixed <- df[df$indiv_prey_weight != 0, ]
df_fixed <- df_fixed[df_fixed$pred_weight_g != 0, ]
df_fixed <- df_fixed[df_fixed$indiv_prey_weight != Inf, ]

#a smaller (random) sample of the overall data set
#to check errors:
#anyNA(renamed_df_small$ppmr)
#any(renamed_df_small$ppmr == 0)

renamed_df = df_fixed %>% 
  mutate(pred_species = replace(pred_species, pred_species == "Clupea harengus", "Herring")) %>% 
  mutate(pred_species = replace(pred_species, pred_species == "Sprattus sprattus", "Sprat")) %>%
  mutate(pred_species = replace(pred_species, pred_species == "Gadus morhua", "Cod")) %>% 
  mutate(pred_species = replace(pred_species, pred_species == "Melanogrammus aeglefinus", "Haddock")) %>% 
  mutate(pred_species = replace(pred_species, pred_species == "Merlangius merlangus", "Whiting")) %>%
  mutate(pred_species = replace(pred_species, pred_species == "Micromesistius poutassou", "Blue Whiting")) %>%
  mutate(pred_species = replace(pred_species, pred_species == "Trisopterus esmarkii", "Norway Pout")) %>%
  mutate(pred_species = replace(pred_species, pred_species == "Trisopterus minutus", "Poor Cod")) %>%
  mutate(pred_species = replace(pred_species, pred_species == "Merluccius merluccius", "European Hake")) %>%
  mutate(pred_species = replace(pred_species, pred_species == "Lophius piscatorius", "Monkfish")) %>%
  mutate(pred_species = replace(pred_species, pred_species == "Trachurus trachurus", "Horse Mackerel")) %>%
  mutate(pred_species = replace(pred_species, pred_species == "Scomber scombrus", "Mackerel")) %>%
  mutate(pred_species = replace(pred_species, pred_species == "Limanda limanda", "Common Dab")) %>%
  mutate(pred_species = replace(pred_species, pred_species == "Pleuronectes platessa", "Plaice")) %>%
  mutate(pred_species = replace(pred_species, pred_species == "Lepidorhombus whiffiagonis", "Megrim")) %>%
  mutate(pred_species = replace(pred_species, pred_species == "Solea solea", "Sole")) %>% 
  mutate(pred_species = replace(pred_species, pred_species == "Capros apers", "Boarfish"))

species_list <- list("Herring", "Sprat", "Cod", "Haddock", "Whiting", "Blue Whiting", "Norway Pout", "Poor Cod", "European Hake", "Monkfish", "Horse Mackerel", "Mackerel", "Common Dab", "Plaice", "Megrim", "Sole", "Boarfish")

renamed_df <- renamed_df[renamed_df$pred_species %in% species_list, ]

#to change to bigger df, just remove the bit below
#renamed_df <- renamed_df[1:5000,]
#renamed_df <- renamed_df[sample(nrow(df), 5000), ]
```

# Distribution of prey type eaten for each predator
```{r distribution of prey, echo = TRUE}
#species_list includes Boarfish which doesn't have values, so add 'length()-1' to account for this - i.e. doesn't analyse any Boarfish data because it's nonexistent

prey_type_list <- list("benthos", "fish", "nekton", "other", "zooplankton")

for (i in 1:(length(species_list)-1)){
  b=f=n=o=z=0
  first_species <- renamed_df %>% filter(renamed_df$pred_species == fixed(species_list[i]))
  
  benthos <- first_species[first_species$prey_type == fixed("benthos"),]
  b <- length(benthos$haul_id)
    
  fish <- first_species[first_species$prey_type == fixed("fish"),]    
  f <- length(fish$haul_id)
    
  nekton <- first_species[first_species$prey_type == fixed("nekton"),]
  n <- length(nekton$haul_id)
    
  other <- first_species[first_species$prey_type == fixed("other"),]
  o <- length(other$haul_id)
    
  zoo <- first_species[first_species$prey_type == fixed("zooplankton"),]
  z <- length(zoo$haul_id)

  pie(c(b,f,n,o,z), prey_type_list, main = species_list[i])
}

```

These are individual pie charts showing the distribution of the type of prey each predator eats.



# Ave PPMR for individual species, weighted by prey biomass:
```{r ave PPMR biomass, echo = TRUE}
#Separated into individual plots for each predator species -> using facet_wrap for the variable (pred_species)
ggplot(data = renamed_df, aes(x=log(ppmr)), group=1) + 
  labs(title = "log(PPMR) v. biomass density of prey", x="log(PPMR)", y="Biomass density of prey") +
  facet_wrap(~renamed_df$pred_species, scale="free_y") + 
  theme(strip.text = element_text(size = 5)) +
  geom_density(aes(weight = prey_weight_g), colour="red")

#poor cod and boarfish?

for (i in 1:length(species_list)){
  name <- species_list[i]
  first_species <- renamed_df %>% 
    filter(pred_species == fixed(name))
  ggplot(data = first_species, aes(x=log(ppmr)), group=1) + 
          labs(title = name, x="log(PPMR)", y="biomass density of prey") +
          geom_density(aes(weight = prey_weight_g), colour="red") + 
          theme(plot.title = element_text(size=15))
  av <- weighted.mean(first_species$ppmr, w = first_species$prey_weight_g, na.rm = TRUE)
  stan_dev <- sqrt(wtd.var(first_species$ppmr, w = first_speciesF$prey_weight_g, na.rm = TRUE))
  #make standard deviation weighted by prey biomass
  print(paste(name, av, stan_dev))
}
``` 
Looking for the most common PPMR for each individual species.

A graph of the weighted ppmr for each species against the biomass density of the prey.
Prints the mean ppmr, as weighted by prey biomass.

# Ave PPMR for individual species, weighted by number of prey:
```{r ave PPMR prey no., echo = TRUE}
#Separated into individual plots for each predator species -> using facet_wrap for the variable (pred_species)
ggplot(data = renamed_df, aes(x=log(ppmr)), group=1) + 
  labs(title = "Scatter plot: log(PPMR) v. number density of prey", x="log(PPMR)", y="Number density of prey") +
  facet_wrap(~renamed_df$pred_species, scale="free_y") + 
  theme(strip.text = element_text(size = 5)) +
  geom_density(aes(weight = no._prey_per_stmch), colour="red")

for (i in 1:length(species_list)){
  name <- species_list[i]
  first_species <- renamed_df %>% 
    filter(pred_species == fixed(name))
  ggplot(data = first_species, aes(x=log(ppmr)), group=1) + 
          labs(title = name, x="log(PPMR)", y="Number density of prey") +
          geom_density(aes(weight = no._prey_per_stmch), colour="red") + 
          theme(plot.title = element_text(size=15))
  av <- weighted.mean(first_species$ppmr, w = first_species$no._prey_per_stmch, na.rm = TRUE)
  stan_dev <- sqrt(wtd.var(first_species$ppmr, w = first_speciesF$no._prey_per_stmch, na.rm = TRUE))
  print(paste(name, av, stan_dev))
}
```
Looking for the most common PPMR for each individual species.

A graph of the weighted ppmr for each species against the number density of the prey.
Prints the mean ppmr, as weighted by number of prey.


# Specific PPMR calculations by different weightings for Herring species
```{r Herring PPMR, echo = TRUE} 
first_species <- renamed_df %>% filter(pred_species == fixed("Herring"))
#separate data set of a single species

renamed_df = renamed_df %>% mutate(lppmr = log(ppmr))
#adding a column of log(ppmr) values

herringDF <- renamed_df %>% 
    filter(pred_species == fixed("Herring"))

herringmean = mean(herringDF$lppmr)
herringSD = sqrt(var(herringDF$lppmr))
#these are non-weighted means
#var is variance, i.e. ave. distance from each point to the mean
#mean is the arithmetic mean of log(ppmr)

bio_herringmean = weighted.mean(herringDF$lppmr, w = herringDF$prey_weight_g, na.rm = TRUE)
bio_herringSD = sqrt(wtd.var(herringDF$lppmr, w = herringDF$prey_weight_g, na.rm = TRUE))
#mean and standard deviation, weighted by the prey biomass


ggplot(data = first_species, aes(x=log(ppmr)), group=1) + 
          labs(title = "Herring, weighted by prey biomass 
               (with a normal distribution, weighted by prey biomass)", x="log(PPMR)",y="Biomass density of observations") +
          geom_density(aes(weight = prey_weight_g), colour="red") + 
          theme(plot.title = element_text(size=15)) + 
          stat_function(fun = dnorm, args= with(herringDF, c(mean = bio_herringmean, sd = bio_herringSD)))

no_herringmean = weighted.mean(herringDF$lppmr, w = herringDF$no._prey_per_stmch, na.rm = TRUE)
no_herringSD = sqrt(wtd.var(herringDF$lppmr, w = herringDF$no._prey_per_stmch, na.rm = TRUE))
#weighted standard deviation found using the weighted variance
#mean and standard deviation, weighted by the number of prey

ggplot(data = first_species, aes(x=log(ppmr)), group=1) + 
          labs(title = "Herring, weighted by no. of prey 
               (with a normal distribution, weighted by prey biomass)", x="log(PPMR)", y="No. density of observations") +
          geom_density(aes(weight = no._prey_per_stmch), colour="red") + 
          theme(plot.title = element_text(size=15)) +
          stat_function(fun = dnorm, args= with(herringDF, c(mean = no_herringmean, sd = no_herringSD)))
```

Plotting the distribution of Herring log(PPMR) as weighted by prey biomass and number of prey (with weighted normal distribution curves plotted over the top of each).

Prey biomass weighted mean: 6.629177
No. of prey weighted mean: 13.54102

The mean is shifted by 6.911843.

What does this mean?

# prey weight v. number density of prey
```{r indiv_prey_weight v. no. density of prey, echo = TRUE}
ggplot(data = renamed_df, aes(indiv_prey_weight, no._prey_per_stmch)) + 
  labs(title = "prey weight v. number of prey per stomach", x="prey weight (g)", y="No. of prey per predator stomach") + 
  geom_point(size=0.75)

ggplot(data = renamed_df, aes(log(indiv_prey_weight), no._prey_per_stmch)) + 
  labs(title = "log(prey weight) v. number of prey per predator stomach", x="log(prey weight)", y="No. of prey") + 
  geom_point(size=0.5)
#Some interesting results --> introduce colours to show ships  

renamed_df$'haul_id_short' <- gsub("\\-.*", "", renamed_df$'haul_id')
renamed_df$'haul_id_short' <- gsub("_", "", renamed_df$'haul_id_short')
#rename haul_id values -> separate by ship names (e.g. CLYDE) rather than complete id (e.g. CLYDE-1935-6)

haul_list <- unique(renamed_df$haul_id_short)

interesting_haul <- filter(renamed_df, haul_id_short=='CLYDE'|haul_id_short=='END04'|haul_id_short=='CORY08'|haul_id_short=='LUC'|haul_id_short=='HIDDINK'|haul_id_short=='EXCmacDATSTO815'|haul_id_short=="Excmacdatsto815error")

haul_list_interesting <- unique(interesting_haul$haul_id_short)

#Vertical lines: CLYDE, BLEGVARD, CIROL05, END03, LAST, LUC (*), DUBUIT, HARDY, JOHNSON

#Curves: END04

#Horizontal lines: CORY08, HIDDINK

#Identical: EXCmacDATSTO815, Excmacdatsto815error

#log(prey weight) v. number of prey per stomach - separated by ship names
for (i in 1:length(haul_list_interesting)){
  haul_df <- interesting_haul %>% filter(interesting_haul$haul_id_short == fixed(haul_list_interesting[i]))
  print(ggplot (data = haul_df, aes(x=log(indiv_prey_weight), y=no._prey_per_stmch)) + 
  labs(title = haul_list_interesting[i], x="log(prey weight)", y="No. prey per stomach") + 
  geom_point(size=.1, colour="red") +
  theme(strip.text = element_text(size = 5)))
}

ggplot (data = interesting_haul, aes(x=log(indiv_prey_weight), y=no._prey_per_stmch)) + 
  labs(title = "hello", x="log(prey weight)", y="No. prey per stomach") + 
  geom_point(size=.1, colour="red") +
  theme(strip.text = element_text(size = 5)) + facet_wrap(~interesting_haul$haul_id_short, scale="free_y")

#create new facet_wrap() over: vertical lines, curves, horizontal lines, identical

```

Playing around with data to see any specific correlations; what is the distribution of the weight of prey recorded.

1. Prey weight v. no. prey per stomach
2. Log (prey weight) v. no. prey per stomach -> showed some interesting results, so added colours to identify individual ships
3. Graph 3., but with points from each ship plotted on separate graphs -> note y prop. to e^-x relation for END04; lots of observations for single weights for LUC; lots of the same no. of fish observations for CORY08 and HIDDINK.

Also, I found that the EXCmacDATSTO815 and Excmacdatsto815error gave exactly the same data (which might need to be considered in later calculations).


# prey weight v. pred weight

```{r prey weight v. pred weight, echo = TRUE}
ggplot (data = renamed_df, aes(indiv_prey_weight, pred_weight_g)) +
  geom_point(size=0.5) + 
  labs(title = "Predator v. prey mass plot", x="Prey mass (g)", y="Predator mass (g)")
#mass since measured in g

ggplot(data = renamed_df, aes(log(indiv_prey_weight), log(pred_weight_g))) + 
  geom_point(size=0.5) + 
  labs(title = "log(Predator mass) v. log(prey mass) plot", x="log(Prey mass)", y="log(Predator mass)") + 
  stat_smooth (method='lm', se=FALSE)

#slope <- coef(lm(log(renamed_df$pred_weight_g)~log(renamed_df$indiv_prey_weight)))
#print(paste("slope of the log(pred) v. log(prey) line of best fit:", slope))
#second part is intercept -> how to separate?

ggplot(data = renamed_df, aes(log(indiv_prey_weight), log(pred_weight_g))) + 
  labs(title = "log(pred. mass) v. log(prey mass) separated by predator species", x="log(prey mass)", y="log(pred. mass)") + 
  geom_point(size=0.2, colour="red") + 
  facet_wrap(~pred_species, scale="free_y") + 
  theme(strip.text = element_text(size = 10))
```

1. Prey weight v. predator weight -> attempting to find a link between the predator mass and the prey mass  
2. log(prey weight) v. log(pred. weight) ->  using log() to see proportionality of the axes, slope of added line should = PPMR  
3. Looking to find correlation between the masses for each individual species; the slope should intersect the y-axis at 0, else our idea for PPMR calculation (pred mass is prop. to prey mass) is incorrect  

# pred weight v. ppmr
```{r pred weight v. ppmr, echo = TRUE}
ggplot(data=renamed_df, aes(log(pred_weight_g), log(ppmr))) + 
  geom_point(size=0.5) + 
  labs(title = "log(pred mass) v. log(ppmr) plot", x="log(Pred mass)", y="log(PPMR)") + 
  stat_smooth (method='lm', se=FALSE)

slope2 <- coef(lm(log(renamed_df$ppmr)~log(renamed_df$pred_weight_g)))
print(paste("slope of the log(ppmr) v. log(pred_weight) line of best fit:", slope2))
#slope of the above plot

ggplot(data=renamed_df, aes(log(pred_weight_g), log(ppmr))) + 
  geom_point(size=0.5) +
  labs(title = "log(pred mass) v. log(ppmr) plot", x="log(Pred mass)", y="log(PPMR)") + stat_smooth (method='lm', se=FALSE) + 
  facet_wrap(~pred_species, scale="free_y") + 
  stat_smooth(method='lm', se=FALSE)

species_df <- renamed_df %>% filter(pred_species == fixed("Poor Cod"))

ggplot(data=species_df, aes(log(pred_weight_g), log(ppmr))) + 
  geom_point(size=0.5) +
  labs(title = "log(pred mass) v. log(ppmr) plot: Poor Cod", x="log(Pred mass)", y="log(PPMR)") + stat_smooth (method='lm', se=FALSE) + 
  facet_wrap(~pred_species, scale="free_y") + 
  stat_smooth(method='lm', se=FALSE)

species_slope <- coef(lm(log(species_df$ppmr)~log(species_df$pred_weight_g)))
print(paste("slope of the log(ppmr) v. log(pred_weight) line of best fit:", species_slope))

#for (i in 1:length(species_list)){
# name <- species_list[i]
 #first_species <- renamed_df %>% filter(pred_species == fixed(name))
 #grad <- coef(lm(log(first_species$ppmr)~log(first_species$pred_weight_g)))
 #print(paste(name, grad)) 
#}
```

log(pred mass) v. log(ppmr) -> is pred. mass related to ppmr? 

We want them to not be proportional (i.e. slope = 0).

CHECK: IS THIS NO. OF POINTS RECORDED OR NO. POINTS*NO. PREY PER STOMACH

```{r first mixed effects, echo=TRUE}

lmer(log(ppmr) ~ pred_species + prey_weight_g + (1|haul_id_short), data = renamed_df)

lmer(log(ppmr) ~ pred_species + prey_weight_g + (1|haul_id_short) + (1|year) + (1|no._prey_per_stmch) + (1|ices_rectangle), data = renamed_df)

#lmer(fixed ~ (1|random) + linear)
```

Fixed effect: log(ppmr)

Random effects: haul_id_short, year

Linear fixed effects: pred_species, prey_weight_g

