---
title: 'Dissertation'
output: html_document
date: "2022-10-27"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning=FALSE) #doesn't show code or error messages in knit
```

```{r library, message=FALSE, include=FALSE} 
#don't print outputs
library(dplyr)
library(tidyverse) 
library(knitr)
library(readr)
library(ggplot2)
```

```{r block, message=FALSE, include=FALSE}
(load("stomach_dataset.Rdata")) #change to access via github repository
#the dataset is already named stom_df
stom_df
df <- stom_df %>% transmute(haul_id, latitude, longitude, year, pred_species, pred_weight_g, pred_length_cm, prey_species, prey_weight_g, prey_type = prey_funcgrp, indiv_prey_weight = prey_ind_weight_g, no._prey_per_stmch = prey_count / n_stomachs, ppmr) 
df
```

```{r cleaning smaller, message=FALSE, include=FALSE}
df_smaller <- df[1:5000,]
#a smaller sample of the overall data set
#to change to bigger df, just rename the bit below

df_fixed <- df_smaller[df_smaller$indiv_prey_weight != 0, ]

#to check errors:
#anyNA(renamed_df_small$ppmr)
#any(renamed_df_small$ppmr == 0)

renamed_df = df_smaller_fixed %>% 
  mutate(pred_species = replace(pred_species, pred_species == "Clupea harengus", "Herring")) %>% 
  mutate(pred_species = replace(pred_species, pred_species == "Sprattus sprattus", "Sprat")) %>%
  mutate(pred_species = replace(pred_species, pred_species == "Gadus morhua", "Cod")) %>% 
  mutate(pred_species = replace(pred_species, pred_species == "Melanogrammus aeglefinus", "Haddock")) %>% 
  mutate(pred_species = replace(pred_species, pred_species == "Merlangius merlangus", "Whiting")) %>%
  mutate(pred_species = replace(pred_species, pred_species == "Micromesistius poutassou", "Blue Whiting")) %>%
  mutate(pred_species = replace(pred_species, pred_species == "Trisopterus esmarkii", "Norway Pout")) %>%
  mutate(pred_species = replace(pred_species, pred_species == "Trisopterus minutus", "Poor Cod")) %>%
  mutate(pred_species = replace(pred_species, pred_species == "Merluccius merluccius", "European Hake")) %>%
  mutate(pred_species = replace(pred_species, pred_species == "Lophius piscatorius", "Monkfish")) %>%
  mutate(pred_species = replace(pred_species, pred_species == "Trachurus trachurus", "Horse Mackerel")) %>%
  mutate(pred_species = replace(pred_species, pred_species == "Scomber scombrus", "Mackerel")) %>%
  mutate(pred_species = replace(pred_species, pred_species == "Limanda limanda", "Common Dab")) %>%
  mutate(pred_species = replace(pred_species, pred_species == "Pleuronectes platessa", "Plaice")) %>%
  mutate(pred_species = replace(pred_species, pred_species == "Lepidorhombus whiffiagonis", "Megrim")) %>%
  mutate(pred_species = replace(pred_species, pred_species == "Solea solea", "Sole")) %>% 
  mutate(pred_species = replace(pred_species, pred_species == "Capros apers", "Boarfish"))


species_list <- list("Herring", "Sprat", "Cod", "Haddock", "Whiting", "Blue Whiting", "Norway Pout", "Poor Cod", "European Hake", "Monkfish", "Horse Mackerel", "Mackerel", "Common Dab", "Plaice", "Megrim", "Sole", "Boarfish")

renamed_df <- renamed_df[ ! renamed_df$pred_species %in% species_list, ]
```

\newpage

##Ave PPMR for individual species, weighted by prey biomass:
```{r ave PPMR}
for (i in 1:length(species_list)){
  name <- species_list[i]
  first_species <- renamed_df %>% 
    filter(pred_species == fixed(name))
  ggplot(data = first_species, aes(x=log(ppmr)), group=1) + 
          labs(title = name, x="log(PPMR)", y="No. density of observations by prey biomass") +
          geom_density(aes(weight = prey_weight_g), colour="red") + 
          theme(plot.title = element_text(size=15))
  av <- weighted.mean(first_species$ppmr, w = first_species$prey_weight_g, na.rm = TRUE)
  stan_dev <- sd(first_species$ppmr, na.rm = TRUE)
  print(paste(name, av, stan_dev))
}

#Separated into individual plots for each species -> using facet_wrap for the variable (pred_species)
ggplot(data = renamed_df, aes(x=log(ppmr), y=no._prey_per_stmch), group=1) + 
  labs(title = "Scatter plot: log(PPMR) v. number density of prey separated by species", x="log(PPMR)", y="No. prey per stomach") +
  geom_density(aes(weight = prey_weight_g), colour="red") + 
  facet_wrap(~new_renamed_df$pred_species, scale="free_y") + 
  theme(strip.text = element_text(size = 5))
``` 
Looking for the most common PPMR for each individual species

#Ave PPMR for individual species, weighted by number of prey:
```{r ave2 PPMR}
for (i in 1:length(species_list)){
  name <- species_list[i]
  first_species <- renamed_df %>% 
    filter(pred_species == fixed(name))
  print(ggplot(data = first_species, aes(x=log(ppmr)), group=1) + 
          labs(title = name, x="log(PPMR)", y="No. density of observations by number") +
          geom_density(aes(weight = no._prey_per_stmch), colour="red") + 
          theme(plot.title = element_text(size=15)))
  av <- weighted.mean(first_species$ppmr, w = first_species$no._prey_per_stmch, na.rm = TRUE)
  stan_dev <- sd(first_species$ppmr,  na.rm = TRUE)
  print(paste(name, av, stan_dev))
}
```

##Specific PPMR calculations by different weightings for Herring species
```{r Herring PPMR} 
first_species <- renamed_df %>% filter(pred_species == fixed("Herring"))
#separate data set of a single species

print(ggplot(data = first_species, aes(x=log(ppmr)), group=1) + 
          labs(title = "Herring, weighted by prey biomass", x="log(PPMR)", y="No. density of observations") +
          geom_density(aes(weight = prey_weight_g), colour="red") + 
          theme(plot.title = element_text(size=15)))
av <- weighted.mean(first_species$ppmr, w = first_species$prey_weight_g, na.rm = TRUE)
stan_dev <- sd(first_species$ppmr, na.rm = TRUE)
print(paste("ave PPMR weighted by prey biomass:", av))
print(paste("standard deviation:", stan_dev))

print(ggplot(data = first_species, aes(x=log(ppmr)), group=1) + 
          labs(title = "Herring, weighted by no. of prey", x="log(PPMR)", y="No. density of observations") +
          geom_density(aes(weight = no._prey_per_stmch), colour="red") + 
          theme(plot.title = element_text(size=15)))
av <- weighted.mean(first_species$ppmr, w = first_species$no._prey_per_stmch, na.rm = TRUE)
stan_dev <- sd(first_species$ppmr,  na.rm = TRUE)
print(paste("ave PPMR weighted by no. of prey:", av))
print(paste("standard deviation:", stan_dev))
```


\newpage

##prey weight v. number density of prey
```{r indiv_prey_weight v. no. density of prey}
ggplot(data = renamed_df, aes(indiv_prey_weight, no._prey_per_stmch)) + labs(title = "prey weight v. number of prey per stomach", x="prey weight (g)", y="No. of prey per predator stomach") + geom_point(size=0.75)

ggplot(data = renamed_df, aes(log(indiv_prey_weight), no._prey_per_stmch)) + labs(title = "log(prey weight) v. number of prey per predator stomach", x="log(prey weight)", y="No. of prey", caption = "Identifying any 'interesting' looking outputs") + geom_point(size=0.5)
#Some interesting results --> introduce colours to show ships  

renamed_df$'haul_id_short' <- gsub("\\-.*", "", renamed_df$'haul_id')
#rename haul_id values -> separate by ship names (e.g. CLYDE) rather than complete id (e.g. CLYDE-1935-6)

#separated by ship names
ggplot (data = renamed_df, aes(x=log(indiv_prey_weight), y=no._prey_per_stmch)) + labs(title = "log(prey weight) v. number of prey per stomach", x="log(prey weight)", y="No. prey per stomach") + geom_point(size=.1, colour="red") + facet_wrap(~haul_id_short, scale="free_y") + theme(strip.text = element_text(size = 5))
```

Playing around with data to see any specific correlations; what is the distribution of the weight of prey recorded.

1. Prey weight v. no. prey per stomach
2. Log (prey weight) v. no. prey per stomach -> showed some interesting results, so added colours to identify individual ships
3. Graph 3., but with points from each ship plotted on separate graphs -> note y prop. to e^-x relation for END04, lots of observations for single weights for LANCELOT; lots of the same no. of fish observations for BULLEN

\newpage

##prey weight v. pred weight

```{r prey weight v. pred weight}
ggplot (data = renamed_df, aes(indiv_prey_weight, pred_weight_g)) + geom_point(size=0.5) + labs(title = "Predator v. prey mass plot", x="Prey mass (g)", y="Predator mass (g)")
#mass since measured in g

ggplot(data = renamed_df, aes(log(indiv_prey_weight), log(pred_weight_g))) + geom_point(size=0.5) + labs(title = "log(Predator mass) v. log(prey mass) plot", x="log(Prey mass)", y="log(Predator mass)") + stat_smooth (method='lm', se=FALSE)

slope <- coef(lm(log(renamed_df$pred_weight_g)~log(renamed_df$indiv_prey_weight)))
print(paste("slope of the log(pred) v. log(prey) line of best fit:", slope))
#second part is intercept -> how to separate?

ggplot(data = renamed_df, aes(log(indiv_prey_weight), log(pred_weight_g))) + labs(title = "log(pred. mass) v. log(prey mass) separated by predator species", x="log(prey mass)", y="log(pred. mass)") + geom_point(size=0.2) + facet_wrap(~pred_species, scale="free_y") + theme(strip.text = element_text(size = 2))
```
1. Prey weight v. predator weight -> attempting to find a link between the predator mass and the prey mass
2. log(prey weight) v. log(pred. weight) ->  using log() to see proportionality of the axes, slope of added line should = PPMR
3. Looking to find correlation between the masses for each individual species; the slope should intersect the y-axis at 0, else our idea for PPMR calculation (pred mass is prop. to prey mass) is incorrect

##pred weight v. ppmr
```{r pred weight v. ppmr}
ggplot(data=renamed_df, aes(log(pred_weight_g), log(ppmr))) + geom_point(size=0.5) + labs(title = "log(pred mass) v. log(ppmr) plot", x="log(Pred mass)", y="log(PPMR)") + stat_smooth (method='lm', se=FALSE)
                                                                                           
#caption = "The slope is not =1, so the PPMR is indept. of pred. mass (as desired)")                      

slope2 <- coef(lm(log(renamed_df$ppmr)~log(renamed_df$pred_weight_g)))
print(paste("slope of the log(ppmr) v. log(pred_weight) line of best fit:", slope2))
#slope of the above plot

#ADD BITS
```
log(pred mass) v. log(ppmr) -> is pred. mass related to ppmr? 

We wanting them to not be proportional (i.e. slope = 0).



##log(PPMR) v. number density of prey

```{r log(PPMR) v. prey no. density, fig.align="center", out.width='100%'}

ggplot(data = renamed_df, aes(log(ppmr), no._prey_per_stmch)) + labs(title = "Scatter plot: log(PPMR) v. no. of prey per stomach", x="log(PPMR)", y="no. prey per stomach", caption = "Trying to find the most common PPMR over all the fish") + geom_point(size=0.5)
#gives the most common PPMR for fish

ggplot(data = renamed_df, aes(x=log(ppmr))) + labs(title = "Histogram: log(PPMR) v. number of observations", x="log(PPMR)", y="No. observations", caption = "no. observation for each value of log(PPMR), i.e. seeing more clearly what the most common PPMR is across all fish species") + geom_histogram(binwidth=.5, colour="black", fill="white")
#Using a histogram to show total no. of observations (rather than scatter plots)

#CHECK: IS THIS NO. OF POINTS RECORDED OR NO. POINTS*NO. PREY PER STOMACH

#Separated into individual plots for each species -> using facet_wrap for the variable (pred_species)
ggplot(data = renamed_df, aes(x=log(ppmr), y=no._prey_per_stmch), group=1) + labs(title = "Scatter plot: log(PPMR) v. number density of prey separated by species", x="log(PPMR)", y="No. prey per stomach", caption = "looking for the most common PPMR for each individual species") + geom_point(size=0.2, colour="red") + facet_wrap(~pred_species, scale="free_y", ncol=3) + theme(strip.text = element_text(size = 5))

#Changed into a density plot
ggplot(data = renamed_df, aes(x=log(ppmr)), group=1) + labs(title = "Histogram: log(PPMR) v. number of observations, separated by species", x="log(PPMR)", y="No. observations", caption = "same idea as above but using a histogram to show most common PPMR per species in a clearer way - the most common should be the one with the single largest bar") + geom_histogram(binwidth=.5, colour="blue", fill="white") + facet_wrap(~pred_species, scale="free_y") + theme(strip.text = element_text(size = 5))

first_species <- renamed_df %>% filter(pred_species == fixed("Herring"))
#separate data set of a single species

ggplot(data = first_species, aes(x=log(ppmr)), group=1) + labs(title = "Histogram: log(PPMR) v. no. of observations: Herring", x="log(PPMR)", y="No. observations") + geom_histogram(binwidth=.5, colour="black", fill="white") + theme(plot.title = element_text(size=15))
```

